---
title: "Registering Incomplete Curves"
author: "Alexander Bauer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Registering Incomplete Curves}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 5
)
```

This vignette outlines the functionalities of the `registr` package with
regard to incomplete curves.

```{r load_libraries, echo = FALSE}
library(registr)
library(ggplot2)
theme_set(theme_minimal() + theme(plot.title = element_text(hjust = 0.5)))
```

## Introduction

Incomplete curves arise in many applications. Incompleteness refers to functional
data where (some) curves were not observed until the very end of the common
domain.
Such a data structure is e.g. observed in the presence of drop-out in panel studies.
Exemplarily, we showcase the following functionalities on data from the Berkeley
Growth Study (see `?growth_incomplete`) where we artificially simulated that
children dropped out of the study at some point in time:

```{r Berkeley data}
dat <- registr::growth_incomplete

# sort the data by the amount of incompleteness
ids <- levels(dat$id)
dat$id <- factor(dat$id, levels = ids[order(sapply(ids, function(curve_id) {
	sum(!is.na(dat$value[dat$id == curve_id]))
}))])

# spaghetti plot
ggplot(dat, aes(x = index, y = value, group = id)) +
	geom_line(alpha = 0.2) +
	xlab("t* [observed]") + ylab("Derivative") +
	ggtitle("First derivative of growth curves")
```


```{r Berkeley data 2, fig.height = 4}
# lasagna plot
ggplot(dat, aes(x = index, y = id, col = value)) + 
	geom_line(lwd = 2.5) +
	scale_color_continuous("Derivative", high = "midnightblue", low = "lightskyblue1") +
  xlab("t* [observed]") + ylab("curve") +
  ggtitle("First derivative of growth curves") +
	theme(panel.grid  = element_blank(),
				axis.text.y = element_blank())
```


## Incomplete curve methodology

We adapt the registration methodology outlined in Wrobel et al. (2019) to
handle incomplete curves. Since each curve potentially has an individual range
of its observed time domain, the spline basis for estimating a curve's warping
function is defined individually for each curve, based on a given number of
basis functions.

It often is a quite strict assumption in incomplete data
settings that all warping functions end on the diagonal, i.e. that the individual,
observed part of the whole time domain is not (to some extent) distorted.
Therefore, the `registr` package gives the additional option to estimate
warping functions without the constraint that the endpoint lies on the diagonal.

On the other end, if we fully remove the constraint for the endpoints
to lie on the diagonal, this can lead to very extreme and unrealistic distortions
of the time domain. This problem is further accompanied by the fact that
the assessment of some given warping to be realistic or unrealistic can heavily
vary between different applications.
As of this reason, our method includes a penalization parameter $\lambda$ that
has to be set manually to specify which kinds of distortions are deemed realistic
in the application at hand.

Mathematically speaking, we add a penalization term to the likelihood $\ell(i)$ 
for curve $i$:
$$
\begin{aligned}
\ell_{pen}(i) &= \ell(i) - \lambda \cdot pen(i), \\
\text{with} \ \ \ 
pen(i) &= \left( \hat{h}^{-1}_i(t^*_{max,i}) - t^*_{max,i} \right)^2,
\end{aligned}
$$
where $t^*_{max,i}$ is the maximum of the observed time domain of curve $i$ and
$\hat{h}^{-1}_i(t^*_{max,i})$ the inverse warping function evaluated at this
maximum.

The higher the penalization parameter $\lambda$, the more the endpoint
of the warping function is forced towards an identity warp, i.e. the diagonal line.
Given a specific application, $\lambda$ should be chosen s.t.
unrealistic distortions of the time domain are prevented.
To do so, the user has to run the registration approach multiple times with
different $\lambda$'s to find an optimal value.


## Application on incomplete growth data

By default, both functions `register_fpca` and `registr` include the argument
`preserve_domain = TRUE` to constrain all warping functions to end on the diagonal.

```{r application 1}
reg1 <- registr(Y = dat, family = "gaussian")

ggplot(reg1$Y, aes(x = tstar, y = index, group = id)) + 
	geom_line(alpha = 0.2) +
	xlab("t* [observed]") + ylab("t [registered]") +
	ggtitle("Estimated warping functions")
```

```{r application 1 lasagna, fig.height = 4}
ggplot(reg1$Y, aes(x = index, y = id, col = value)) + 
	geom_line(lwd = 2.5) +
	scale_color_continuous("Derivative", high = "midnightblue", low = "lightskyblue1") +
  xlab("t [registered]") + ylab("curve") +
  ggtitle("Registered curves") +
	theme(panel.grid  = element_blank(),
				axis.text.y = element_blank())
```

The assumption can be dropped by setting `preserve_domain = FALSE` and
some nonnegative value for the penalization parameter `lambda_endpoint`.
The higher `lambda_endpoint` is chosen, the more all warping functions are forced
to end close to the diagonal.

#### Small `lambda_endpoint`

```{r application 2}
reg2 <- registr(Y = dat, family = "gaussian",
								preserve_domain = FALSE, lambda_endpoint = 1)

ggplot(reg2$Y, aes(x = tstar, y = index, group = id)) + 
	geom_line(alpha = 0.2) +
	xlab("t* [observed]") + ylab("t [registered]") +
	ggtitle("Estimated warping functions")
```

```{r application 2 lasagna, fig.height = 4}
ggplot(reg2$Y, aes(x = index, y = id, col = value)) + 
	geom_line(lwd = 2.5) +
	scale_color_continuous("Derivative", high = "midnightblue", low = "lightskyblue1") +
  xlab("t [registered]") + ylab("curve") +
  ggtitle("Registered curves") +
	theme(panel.grid  = element_blank(),
				axis.text.y = element_blank())
```

#### Large `lambda_endpoint`

```{r application 3}
reg3 <- registr(Y = dat, family = "gaussian",
								preserve_domain = FALSE, lambda_endpoint = 10)

ggplot(reg3$Y, aes(x = tstar, y = index, group = id)) + 
	geom_line(alpha = 0.2) +
	xlab("t* [observed]") + ylab("t [registered]") +
	ggtitle("Estimated warping functions")
```

```{r application 3 lasagna, fig.height = 4}
ggplot(reg3$Y, aes(x = index, y = id, col = value)) + 
	geom_line(lwd = 2.5) +
	scale_color_continuous("Derivative", high = "midnightblue", low = "lightskyblue1") +
  xlab("t [registered]") + ylab("curve") +
  ggtitle("Registered curves") +
	theme(panel.grid  = element_blank(),
				axis.text.y = element_blank())
```


## Help files

Documentation for individual functions gives more information on their arguments and return objects, and can be pulled up via the following:

- `?register_fpca`
- `?registr`
