---
title: "registr: a vignette"
author: "Julia Wrobel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{registr: a vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

The `registr` package is for registering, or aligning, exponential family functional data. Right now it works for binary or continuous data.

```{r load_libraries, echo = FALSE}
library(registr)
library(boot)
library(ggplot2)
library(dplyr)
library(gridExtra)
```


## 'registr package'

The main function in the package is `register_fpca()`. It calls a registration function, `registr()`, performs curve alignment and an gfpca function which calculates a template to which curves are registered using generalized functional principal component analysis (gfpca). The function that calculates gfpca depends on the family. For `family = "binomial"` the `bfpca()` function performs this step and for `family = "gaussian"` the `fpca_gauss()` function performs this step. The `register_fpca()` function iterates between the alignment and template calculation steps until curves are registered. 

This vignette provides examples that highlight the functionality of the package. 	


## Simulated data

There are two functions for simulating data included in the package: `simulate_functional_data()` and `simulate_unregistered_curves()`. Both functions simulate functional data; the first is intended for testing the gfpca functions in our package and the second is for demonstrating our registration algorithm.

### Simulate data for gfpca

`simulate_functional_data()` simulates data with a population-level mean and two orthogonal principal components based on sine and cosine functions. Below I generate data using this function:

```{r sim_data1}
fpca_data = simulate_functional_data(I = 100, D = 200)

ls(fpca_data)

head(fpca_data$Y)
```

I simulated data with $I = 100$ subjects and $D = 200$ time points per subject. The resulting object,`fpca_data`, is a list that contains the true population-level mean (`alpha`) and principal components (`psi1` and `psi2`), and a dataframe (`Y`). The dataframe `Y` contains variables `id`, `value`, `index` and `latent_mean`, which is consistent with the format our `registr` software requires. `id` indicates data for a particular subject, `index` is the time grid for that subject, evaluated on $[0,1]$, the `value` variable contains binary observations, and `latent_mean` contains continuous observations used to generate the binary observations for the `value` variable. Note that when `family = "binomial"` we will use the binary `value` variable as the observations for each subject and when `family = "gaussian"` we use the `latent_mean` variable as the outcome. This data is plotted below.

```{r plot1_sim1, fig.width=6, echo = FALSE}

Y = fpca_data$Y
pc_df = data.frame(pop_mean = fpca_data$alpha, 
									 psi1 = fpca_data$psi1,
									 psi2 = fpca_data$psi2,
									 index = seq(0, 1, length.out = 200),
									 id = 1)

latent_means = ggplot(Y, aes(index, latent_mean, group = id)) + theme_bw() +
	geom_line(alpha = 0.25) + geom_line(data = pc_df, aes(y = pop_mean), color = "red") 

psi1 = ggplot(pc_df, aes(index, psi1)) + theme_bw() + geom_line(color = "blue") 
psi2 = ggplot(pc_df, aes(index, psi2)) + theme_bw() + geom_line(color = "blue") 

grid.arrange(latent_means, psi1, psi2, ncol = 3)

```

The left panel of the figure above shows the latent means for each subject, along with the population-level mean,  $\alpha(t)$, in red. The middle and right panels show the first and second principal components, $\psi_1(t)$ and $\psi_2(t)$, respectively. Using the $\logit^{-1}$ function we can convert the subject-specific means to probabilities; these probabilities are used to generate the binary values. Binary values and latent probability for one subject in the dataset is shown below.

```{r plot2_sim1, echo = FALSE}
Y %>%
	filter(id == 7) %>%
	ggplot(aes(index, value)) + theme_bw() +
	geom_point(alpha = 0.75, size = 0.25) + geom_line(aes(y = inv.logit(latent_mean))) +
	labs(y = "Pr(Y = 1)")

```

We can alter the score variance for the principal components using the arguments `lambda1` and `lambda2`. The default setting is for all subjects to have the same number of time points. However, by specifying `vary_D = TRUE`, we can generate data with uneven grid lengths for each subject.

### Simulate data for registration

The second function for simulating data, `simulate_unregistered_curves()`, generates curves intended to mimic accelerometer data. Use it to test out the `registr()` and `register_fpca()` functions.

Below I generate data using this function:

```{r sim_data2}
registration_data = simulate_unregistered_curves(I = 100, D = 200)

head(registration_data)

```

I simulated data with $I = 100$ subjects and $D = 200$ time points per subject. The resulting object,`registration_data`, is a data frame with variables `id`, `value`, `index`, `latent_mean`, and `t`. Variables `id`, `value`, and `latent_mean` are the same as those from the `fpca_object` data. The variables `index` and `t` are both time grids. Evaluated on the grid `index` the data is unregistered, and on the grid `t` the data is registered. Registered and unregistered curves are plotted below.

```{r plot_sim2, echo = FALSE, fig.width = 6}

unregistered = registration_data %>%
	ggplot(aes(index, inv.logit(latent_mean), group = id)) + theme_bw() + 
	geom_line(alpha = 0.25) + labs(y = "Pr(Y = 1)")

registered = registration_data %>%
	ggplot(aes(t, inv.logit(latent_mean), group = id)) + theme_bw() + 
	geom_line(alpha = 0.25) + labs(y = "Pr(Y = 1)")

grid.arrange(unregistered, registered, ncol = 2)
```

Each curve has one main peak, but the location of that peak is shifted. When curves are registered the peaks align.



## Registration using `register_fpca()`

### binary data

The main functionality is for binary data.

### continuous data


## `bfpca()` function

Show example of running this code separately. Show subjects and their estimated curves. Show true and estimated means and fpcs.

## `registr()` function

Additional options for registration. Talk about gradient.


## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
